---
title: "Literate Analysis"
author: "Pascale"
format: html
---

## Introduction

### About this data
This data set includes measurements of seawater chemistry along with supporting conductivity, temperature, and depth (CTD) data from the Arctic Ocean near the North Pole. Measurements were taken from sea ice platforms each April or May from 2000-2014. These data were access from the [Artic Data Center](https://orcid.org/0000-0002-3409-8881) on 6/26/2023.

### Setup
```{r}
library(readr)


```

### Read in data
```{r}
#| warning: false
#| message: false
bg_chem <- read_csv("data/BGchem2008data.csv")
```
Exploring data
```{r}
#| eval: false
#| echo: fenced
str(bg_chem)

colnames(bg_chem)

head(bg_chem)

summary(bg_chem)

unique(bg_chem$Date)
```


## Analysis 

### Calculate summary statisitics
```{r}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)


```

### Calculate mean Redfield ratio
```{r}
ratio <- (nitrate + nitrite + amm)/phos
```


### Plot Redfield ratio
```{r}
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4)
```

## Conclusion 

The Redfield ratio for this data is approximately `r round(ratio)`